---
title: "STATS 531 Final Project"
output: html_document
date: "2025-04-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(timeSeriesDataSets)
```

# Dataset 
- Period: 2024-04-06 - 2025-04-05
- CNN Fear & Greed Index
- Bitcoin Close price

```{r}
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("httr")
install.packages("jsonlite")
library(httr)
library(jsonlite)

# Call the API
response <- GET("https://api.alternative.me/fng/?limit=1000")

# Parse JSON content
content_json <- content(response, as = "text", encoding = "UTF-8")
fng_data <- fromJSON(content_json)

# Convert the 'data' field to a data frame
fng_df <- fng_data$data
head(fng_df)

fng_df$date <- as.POSIXct(as.numeric(fng_df$timestamp), origin = "1970-01-01", tz = "UTC")
fng_df$value <- as.numeric(fng_df$value)

fng_final <- fng_df[, c("date", "value")]
names(fng_final) <- c("time", "FG")  # Rename for consistency
head(fng_final)

library(ggplot2)

ggplot(fng_final, aes(x = time, y = FG)) +
  geom_line(color = "steelblue") +
  labs(
    title = "CNN Fear & Greed Index Over Time",
    x = "Date",
    y = "Fear & Greed Index"
  ) +
  theme_minimal()

nrow(fng_final)
```
 
```{r}

# Define the start and end dates as POSIXct objects
start_date <- as.POSIXct("2024-04-06", tz = "UTC")
end_date   <- as.POSIXct("2025-04-06", tz = "UTC")

# Subset the data
fng_subset <- fng_final[fng_final$time >= start_date & fng_final$time <= end_date, ]

# Check the first few rows of the subset
head(fng_subset)

# Plot the subsetted data
library(ggplot2)
ggplot(fng_subset, aes(x = time, y = FG)) +
  geom_line(color = "steelblue") +
  labs(
    title = "CNN Fear & Greed Index (2024-04-06 to 2025-04-06)",
    x = "Date",
    y = "Fear & Greed Index"
  ) +
  theme_minimal()

```

After pre-processing, we have "merged_df" which contains date, FG index and bitcoin daily close prices ranging from 04/06/2024 to 04/05/2025. 

```{r}
# Read the bitcoin dataset (ensure strings are not converted to factors)
bitcoin_ts <- read.csv('bitcoin_2024-04-06_2025-04-06.csv', stringsAsFactors = FALSE)

# Convert the 'Start' column to a POSIXct date object 
bitcoin_ts$Start <- as.POSIXct(bitcoin_ts$Start, format = "%Y-%m-%d", tz = "UTC")

# Merge the F&G index dataset (fng_final) with the bitcoin dataset on matching dates
merged_df <- merge(fng_final, bitcoin_ts, by.x = "time", by.y = "Start")

# Select only the required columns and rename them:
# "date" for the matching date, "fg_index" for the Fear & Greed Index, and "bitcoin_close" for the Bitcoin closing price.
merged_df <- merged_df[, c("time", "FG", "Close")]
names(merged_df) <- c("date", "fg_index", "bitcoin_close")

# Check the first few rows of the merged data
head(merged_df)

```
 
 
```{r}
# First plot the FG Index with its y-axis (left)
plot(merged_df$date, merged_df$fg_index, type = "l", col = "blue",
     xlab = "Date", ylab = "FG Index", main = "FG Index and Bitcoin Price Over Time")

# Overlay the Bitcoin closing price with a new plot
par(new = TRUE)
plot(merged_df$date, merged_df$bitcoin_close, type = "l", col = "red", 
     axes = FALSE, xlab = "", ylab = "")
# Add a right-side y-axis for Bitcoin
axis(side = 4, col = "red", col.axis = "red")
mtext("Bitcoin Close", side = 4, line = 3, col = "red")

# Add a legend
legend("topleft", legend = c("FG Index", "Bitcoin Close"), 
       col = c("blue", "red"), lty = 1)


```
```{r}
set.seed(1)

# Assuming 'merged_df' is already available from your earlier code
# with columns: date, fg_index, bitcoin_close
# For consistency with your ETH example, rename columns to "Date" and "Price"
bitcoin <- merged_df
names(bitcoin) <- c("Date", "FG", "Price")

# Convert the Date column to Date type (assuming it's in "YYYY-MM-DD" format)
bitcoin$Date <- as.Date(bitcoin$Date)

# Order the data by Date (if not already sorted)
bitcoin <- bitcoin[order(bitcoin$Date), ]
head(bitcoin)
#
```
### Bitcoin Price Decomposition: Trend, Noise, Cycles
```{r}
library(zoo)

# Convert Date to numeric for LOESS smoothing
bitcoin_num <- as.numeric(bitcoin$Date)

# Build each component as a zoo object
B_btc    <- zoo(bitcoin$Price, bitcoin$Date)
B_trend  <- zoo(loess(bitcoin$Price ~ bitcoin_num, span = 0.6)$fitted, bitcoin$Date)
B_noise  <- zoo(bitcoin$Price - loess(bitcoin$Price ~ bitcoin_num, span = 0.07)$fitted, bitcoin$Date)
B_cycles <- B_btc - B_trend - B_noise

# Set up 4-row layout for plots
par(mfrow = c(4, 1), mar = c(3, 4, 2, 2), oma = c(1, 1, 5, 1)) # Adjust plot margins

# 1. Original Bitcoin Price
plot(B_btc, main = "Bitcoin Price (Original)", xlab = "Date", ylab = "Price", col = "black")

# 2. Trend component
plot(B_trend, main = "Trend (LOESS, span = 0.6)", xlab = "Date", ylab = "Trend", col = "red")

# 3. Noise component
plot(B_noise, main = "Noise (LOESS Residual, span = 0.07)", xlab = "Date", ylab = "Noise", col = "blue")

# 4. Cyclical component
plot(B_cycles, main = "Cyclical Component", xlab = "Date", ylab = "Cycles", col = "green")
mtext("Bitcoin Price Decomposition: Trend, Noise, Cycles", outer = TRUE, cex = 1.5, font = 2)
```

### FG index Decomposition: Trend, Noise, Cycles
```{r}
FG <- bitcoin$FG   # FG index from merged_df
FG_date <- bitcoin$Date
# Convert Date to numeric for LOESS smoothing
fg_num <- as.numeric(FG_date)

# Build zoo objects
FG_index    <- zoo(FG, FG_date)
FG_trend    <- zoo(loess(FG ~ fg_num, span = 0.6)$fitted, FG_date)
FG_noise    <- zoo(FG - loess(FG ~ fg_num, span = 0.07)$fitted, FG_date)
FG_cycles   <- FG_index - FG_trend - FG_noise

# Plot layout: 4 stacked subplots + outer title
par(mfrow = c(4, 1), mar = c(3, 4, 2, 2), oma = c(1, 1, 5, 1))

plot(FG_index, main = "FG Index (Original)", xlab = "Date", ylab = "Index", col = "black")
plot(FG_trend, main = "Trend (LOESS, span = 0.6)", xlab = "Date", ylab = "Trend", col = "red")
plot(FG_noise, main = "Noise (LOESS Residual, span = 0.07)", xlab = "Date", ylab = "Noise", col = "blue")
plot(FG_cycles, main = "Cyclical Component", xlab = "Date", ylab = "Cycles", col = "green")

# Add overall title
mtext("Fear & Greed Index Decomposition: Trend, Noise, Cycles", 
      outer = TRUE, cex = 1.5, font = 2, line = 2)
```

### Smoothed Periodogram 
```{r}
# Bitcoin Price (from merged_df / bitcoin$Price)
B_ind <- bitcoin$Price

spectrum(B_ind,
         spans = c(20, 20),
         xlab = "Frequency - Cycles per Day",
         main = "Bitcoin Daily Price - Smoothed Periodogram")

# FG Index (from merged_df / bitcoin$FG)
FG_ind <- bitcoin$FG

spectrum(FG_ind,
         spans = c(20, 20),
         xlab = "Frequency - Cycles per Day",
         main = "Fear & Greed Index - Smoothed Periodogram")
```

### Differencing and Log-transform
```{r}
# Compute the continuously compounded log returns and demean them
log_returns <- diff(log(bitcoin$Price))
logd <- log_returns - mean(log_returns)

# Plot the Bitcoin Price over time
par(mai=c(0.8,0.8,0.1,0.1))
plot(bitcoin$Price ~ bitcoin$Date, type="l",
     xlab="Years", ylab="Bitcoin ($)",
     main="Bitcoin Price Over Time")

# Plot the logarithmic transformation of Bitcoin Price
plot(log(bitcoin$Price) ~ bitcoin$Date, type="l",
     xlab="Years", ylab="Log(Bitcoin Price)",
     main="Log-Transformed Bitcoin Price")

# Plot the demeaned log returns of Bitcoin
# (Note: diff reduces the length by one, so use Date[-1])
plot(bitcoin$Date[-1], logd, type="l",
     xlab="Years", ylab="Demeaned Log Return",
     main="Demeaned Log Returns of Bitcoin")
 
```

### ACF of Bitcoin Price
```{r}
# Recalculate if needed
bitcoin$log_price <- log(bitcoin$Price)
log_returns <- diff(bitcoin$log_price)
log_returns_demeaned <- log_returns - mean(log_returns)


# ACF 1: Original price
acf(bitcoin$Price, lag.max = 50, main = "ACF: Original Price")

# ACF 2: Log(price)
acf(bitcoin$log_price, lag.max = 50, main = "ACF: Log(Price)")

# ACF 3: Differenced log(price)
acf(log_returns, lag.max = 50, main = "ACF: Diff Log(Price)")

# ACF 4: Demeaned log returns
acf(log_returns_demeaned, lag.max = 50, main = "ACF: Demeaned Log Returns")

# Shared title
mtext("Figure X. ACF of Bitcoin Price: Original, Log, Differenced, Demeaned",
      outer = TRUE, cex = 1.2, font = 2, line = 1)
```

### ACF of FG Index
```{r}
# Prepare FG Index (from merged_df / bitcoin$FG)
FG_index <- bitcoin$FG
FG_date <- bitcoin$Date

# Log transform
log_FG <- log(FG_index)

# First difference of log
diff_log_FG <- diff(log_FG)

acf(FG_index, lag.max = 50, main = "ACF of Fear & Greed Index (Original)")
acf(log_FG, lag.max = 50, main = "ACF of Log(Fear & Greed Index)")
acf(diff_log_FG, lag.max = 50, main = "ACF of Differenced Log(FG Index)")
```

# Simple Stochastic Volatility Model
### Local Search
```{r}
library(pomp)

# Precompute differenced log returns
bitcoin$log_price <- log(bitcoin$Price)
log_returns <- diff(bitcoin$log_price)
btc <- bitcoin[-1, ]  # Remove first row (NA from diff)
btc$log_return <- log_returns
btc$time <- 1:nrow(btc)
```

```{r}
btc_statenames <- c("V")
btc_paramnames <- c("sigma_omega", "phi", "theta", "V_0")

btc_rproc <- "
  double omega;
  omega = rnorm(0, sigma_omega);
  V = theta * (1 - phi) + phi * sqrt(V) + sqrt(V) * omega;
  if (V < 0) V = 0;
"

btc_rinit <- "V = V_0;"
btc_rmeasure <- "y = rnorm(0, sqrt(V));"
btc_dmeasure <- "lik = dnorm(y, 0, sqrt(V), give_log);"

btc_partrans <- parameter_trans(
  log = c("sigma_omega", "theta", "V_0"),
  logit = "phi"
)

btc_params <- c(
  sigma_omega = 0.001,
  phi = 0.2,
  theta = 0.0004,
  V_0 = 0.002
)

btc.pomp <- pomp(
  data = data.frame(time = btc$time, y = btc$log_return),
  times = "time",
  t0 = 0,
  statenames = btc_statenames,
  paramnames = btc_paramnames,
  rprocess = discrete_time(step.fun = Csnippet(btc_rproc), delta.t = 1),
  rinit = Csnippet(btc_rinit),
  rmeasure = Csnippet(btc_rmeasure),
  dmeasure = Csnippet(btc_dmeasure),
  params = btc_params,
  partrans = btc_partrans
)
```

```{r}
btc_sim <- simulate(btc.pomp, seed = 1)
plot(btc_sim, main = "Simulated vs. Observed Log Returns (Bitcoin)")
```

```{r}
library(doParallel)
library(doRNG)
registerDoParallel(detectCores())
registerDoRNG(1234)

# Tuning
run_level <- 2
Np <- switch(run_level, 100, 1000, 2000)
Nmif <- switch(run_level, 10, 100, 200)
Nreps_eval <- switch(run_level, 4, 10, 20)
Nreps_local <- switch(run_level, 5, 10, 20)

btc_rw_sd <- rw_sd(
  sigma_omega = 0.01,
  phi         = 0.01,
  theta       = 0.01,
  V_0         = ivp(0.01)
)

btc_mif <- foreach(i = 1:Nreps_local, .packages = 'pomp', .combine = c) %dopar% {
  mif2(btc.pomp,
       params = btc_params,
       Np = Np,
       Nmif = Nmif,
       cooling.fraction.50 = 0.5,
       rw.sd = btc_rw_sd)
}

btc_L <- foreach(i = 1:Nreps_local, .packages = 'pomp', .combine = rbind) %dopar% {
  logmeanexp(replicate(Nreps_eval, logLik(pfilter(btc.pomp,
                                                  params = coef(btc_mif[[i]]),
                                                  Np = Np))),
             se = TRUE)
}

btc_results <- data.frame(
  logLik = btc_L[, 1],
  logLik_se = btc_L[, 2],
  t(sapply(btc_mif, coef))
)

summary(btc_results$logLik)
plot(btc_mif)
pairs(~logLik + sigma_omega + theta + phi, data = btc_results)
```

### Global Search
```{r}
btc_box <- rbind(
  V_0 = c(0.0001, 0.02),
  sigma_omega = c(0.0001, 0.1),
  phi = c(0.01, 0.99),
  theta = c(0.0001, 0.02)
)
Nreps_global <- switch(run_level, 10, 20, 100)

btc_global_mif <- foreach(i = 1:Nreps_global, .packages = 'pomp', .combine = c) %dopar% {
  start_params <- apply(btc_box, 1, function(x) runif(1, min = x[1], max = x[2]))
  
  mif2(
    btc_mif[[1]],  # use the model structure from local search
    params = start_params,
    Np = Np,
    Nmif = Nmif,
    cooling.fraction.50 = 0.5,
    rw.sd = btc_rw_sd
  )
}
btc_L_global <- foreach(i = 1:Nreps_global, .packages = 'pomp', .combine = rbind) %dopar% {
  logmeanexp(
    replicate(Nreps_eval, logLik(pfilter(
      btc.pomp,
      params = coef(btc_global_mif[[i]]),
      Np = Np
    ))),
    se = TRUE
  )
}
btc_results_global <- data.frame(
  logLik = btc_L_global[, 1],
  logLik_se = btc_L_global[, 2],
  t(sapply(btc_global_mif, coef))
)

summary(btc_results_global$logLik)

# Save results if desired
write.csv(btc_results_global, "btc_global_params.csv", row.names = FALSE)

# Diagnostic plots
plot(btc_global_mif)
pairs(~logLik + sigma_omega + theta + phi, data = btc_results_global)
```
